% (C) Copyright Vesa Karvonen 2004.
%
% Distributed under the Boost Software License, Version 1.0.
% (See accompanying file LICENSE.)

\documentclass[10pt,a4paper]{report}

\usepackage{graphicx,url}

\begin{document}

%% Setup %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newtheorem{exercise}{Exercise}[section]
\newcommand{\code}{\url}

\def\ebnf#1{
  \begin{tabular}{rclr}
    \input{#1.ltx}
  \end{tabular}
}

%% Title %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\pagestyle{headings}
\pagenumbering{roman}

\author{Vesa~Karvonen}
\title{The Order Metalanguage for C Preprocessor Metaprogramming\\
  \bigskip
  WORKING DRAFT}

\maketitle

%% Copyright %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage

{\footnotesize \raggedright

\copyright{} Copyright Vesa Karvonen 2004.

\smallskip

Distributed under the Boost Software License, Version 1.0.

\bigskip

{\large Boost Software License - Version 1.0 - August 17th, 2003}

\medskip

Permission is hereby granted, free of charge, to any person or
organization obtaining a copy of the software and accompanying
documentation covered by this license (the ``Software'') to use,
reproduce, display, distribute, execute, and transmit the Software,
and to prepare derivative works of the Software, and to permit
third-parties to whom the Software is furnished to do so, all
subject to the following:

\smallskip

The copyright notices in the Software and this entire statement,
including the above license grant, this restriction and the
following disclaimer, must be included in all copies of the
Software, in whole or in part, and all derivative works of the
Software, unless such copies or derivative works are solely in the
form of machine-executable object code generated by a source
language processor.

\smallskip

THE SOFTWARE IS PROVIDED ``AS IS'', WITHOUT WARRANTY OF ANY KIND,
EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE, TITLE AND
NON-INFRINGEMENT. IN NO EVENT SHALL THE COPYRIGHT HOLDERS OR ANYONE
DISTRIBUTING THE SOFTWARE BE LIABLE FOR ANY DAMAGES OR OTHER
LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT
OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS
IN THE SOFTWARE.}

%% Preface %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

Order grew out of my interest in the design of programming languages
and metaprogramming as well as desire to make something original,
fascinating and beautiful. My work on the Boost.Preprocessor library
\cite{boost.pp} had left me unsatisfied. While it was clear to me
that it could theoretically be used to implement \emph{useful}
non-trivial metaprograms, it would have been quite tedious work,
full of nasty little details and limitations to trip you up. I
considered the prospects for further enhancements to the library,
but it didn't seem possible to provide a programming model that I
would have considered pleasant to think in. It occured to me that
instead of attempting to provide semimodular enhancements on top of
the C preprocessor in a bottom-up fashion, it might be possible to
lift the limitations by designing a complete monolithic interpreter
for a high-level source language in a top-down fashion. After some
experiments at interpreting lambda calculus and a couple of
prototype interpreters for early versions of the Order language, I
was left with a feeling that it would be too inefficient to be
practical. The most serious limitation seemed to be the inability of
the early prototypes to generate arbitrary output efficiently. I
decided to forget about the interpreter and move on, but the work
proved too interesting for me to forget completely. Over a long
period of time I found several small incremental improvements to the
interpreter until it suddenly started to look like more than a
theoretical curiosity. Then I knew I just had to finish the work.
Hopefully everyone will be able to find some beauty in the result.
At least I know I've had a lot of fun designing the language.

\subsection*{Intent}

This book introduces the complete Order language and prelude and
presents several complete examples of using the Order language.

While this isn't a formal definition of the Order language, key
constructs of the language will be given approximate\footnote{The C
preprocessor macro replacement mechanism will not be completely
modelled.} formal specifications. The main purpose of presenting
the semantics is to show that the Order language is actually very
much like any ordinary programming language. Also, as this isn't a
book about theoretical aspects of programming languages intended
for students of computer science, but rather the manual of a
programming language intended for practitioners, we will be rather
lax in the definition and use of the formal notations we employ.

While the implementation of the Order interpreter has required
working through a considerable number of technical obstacles, we
will largely ignore such low-level details and concentrate on the
high-level design and use of the language.

\subsection*{Audience}

This book is written for programmers working on C and C++ projects
who would like to use metaprogramming techniques in their work. The
reader should be familiar with the C and C++ programming languages
and have a basic understanding of the C preprocessor. Previous
knowledge of lambda calculus and functional programming will help
to understand the material.

\subsection*{Electronic Resources}

The latest releases of the Order interpreter and documentation can
be found at the Chaos preprocessor SourceForge project:
\url{http://sourceforge.net/projects/chaos-pp}.

\subsection*{Acknowledgements}

I wish to thank Paul Mensonides, who discovered many of the C
preprocessor programming techniques used in the Order interpreter
and the prelude. He also provided the project space for constructing
the Order interpreter and participated in many intense discussions
related to C preprocessor programming. Without Paul's contributions,
the Order interpreter wouldn't exist today.

\bigskip

{\raggedright Vesa Karvonen}

%% Table of Contents %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\tableofcontents

%% Body %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\setcounter{page}{1}
\pagenumbering{arabic}

\chapter{Introduction} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This chapter gives a minimal overview of the main features of the
Order language and interpreter and puts it into perspective as a
metalanguage of the C preprocessor. The last section of this chapter
is an example using the Order interpreter.

\section{What is Order?} %******************************************

Order is a \emph{complete} programming language designed to be
\emph{interpreted} using the \emph{C preprocessor} \cite{c:1999}.
Rather than being a general purpose language, Order is a
\emph{metalanguage} that can be used to \emph{generate} sequences
of \emph{preprocessing tokens}.

Order is a \emph{high-level} \emph{functional} language with
\emph{call-by-value} semantics, providing \emph{first-class
anonymous functions} with \emph{lexically scoped variables} and
supporting \emph{partial application} of functions. The Order
interpreter is implemented in \emph{continuation passing style} and
is fully \emph{tail recursive}. Order is a \emph{reflective}
language providing \emph{first class continuations}, \emph{first
class environments} and an \emph{eval} function. Memory management
in Order is implicit. Order provides \emph{arbitrary-precision
arithmetic} on natural numbers, a comprehensive set of primitive and
\emph{higher-order} functions for an aggregate data type called
\emph{sequence} and \emph{more}.\footnote{If you are not familiar
with some of the terms used here, rest assured, all will become
clear.}

Order is a fairly simple language to learn. It has only a modest
number of syntactic forms with simple semantics and is based on
the \emph{lambda calculus}.\footnote{Programmers familiar with both
Scheme \cite{dybvig:1996,abelson:1996} and SML, in addition to C and
the C preprocessor, should find Order particularly easy to learn.}
Compared to other C preprocessor metaprogramming systems, such as
the Boost Preprocessor library \cite{boost.pp}, Order provides a
clearly higher-level alternative. Rather than providing some
enhancements over the plain C preprocessor designed one problem at a
time, Order implements a complete solution at once.

\section{The C Preprocessor} %**************************************

The C preprocessor (Cpp) is part of both C \cite{c:1999} and C++
\cite{c++:1998} and is mainly used to compose \emph{translation
units} out of \emph{source files} using the \code{#include}
directive and a family of conditional directives (\code{#if},
\code{#ifdef},
\ldots, \code{#endif}). The Cpp is a \emph{low-level}
\emph{macroprocessor} enabling the definition of \emph{object-like}
and \emph{function-like} macros using the \code{#define}
directive.\footnote{The Cpp also has some other features and uses,
but they are not relevant to Order.} The Order interpreter is
implemented using the macro mechanism of the Cpp.

The preprocessor defined by the \cite{c:1999} standard includes a
number of important extensions, namely \emph{variadic} macros and
\emph{placemarkers}, that aren't provided by standard C++. Apart
from those extensions, and a few minor differences, the C and C++
preprocessors are essentially the same and many modern C++
implementations actually provide variadic macros and placemarkers as
an extension as they are likely to be adopted into the next C++
standard.

Bjarne Stroustrup, the inventor of C++, has said that ``I'd like to
see Cpp abolished'' \cite{stroustrup:1994}. Indeed, many consider
the Cpp to be one of the \emph{worst} macroprocessors in history and
for good reasons. We will assume that the reader is all too familiar
with the problems described in \cite{stroustrup:1994} and many other
sources and we will only discuss issues that are directly relevant
to our subject.

The main strength of the Cpp is that it comes with every standards
conforming C and C++ implementation. Compared to external
metaprogramming facilities for C and C++, Cpp ``Works out of the
box.''. This strength should not be underestimated. Library authors,
and programmers in general, would often like to use generative
programming techniques as described in \cite{czarnecki:2000,
graham:1993}. However, code generation with external, non-standard
tools introduces a host of problems typically making the
maintenance, distribution and porting of software much harder.

As implied, the macro mechanism of the Cpp hasn't been designed to
be very powerful. It explicitly disallows recursion by implementing
a \emph{name disabling} scheme, which marks recursively generated
macro names as disabled from macro replacement.\footnote{The include
mechanism can be used recursively, but it is much too low-level a
mechanism to be used for anything non-trivial.} The macro mechanism
also doesn't directly support repetition of any kind. Despite these
limitations, and the fact that the Cpp macro mechanism can't get
into an infinite loop, it has turned out that the Cpp macro
mechanism is, from a practical point of view, \emph{Turing complete}
\cite{lewis:1998}, being capable of evaluating arbitrary algorithms
for any predefined number of computational steps using an evaluator
consisting of a number of macros logarithmic to the number of
computational steps. The practical computational completeness of the
Cpp opens up the possibility of using the Cpp for non-trivial
\emph{code generation}.

Before we continue, let's put the strong claim of practical Turing
completeness into perspective. Let's assume that one would like to
empirically verify the incompleteness of the current Order
interpreter. To do this, one would design a program that
\emph{should}, according to the hypothesis, eventually fail due to
running out of computational steps.\footnote{An infinite loop would
do the trick.} For the sake of getting some concrete numbers, let's
further assume that a Cpp implementation would exists that could
perform 1 billion (1000000000) macro replacements per second, which
should be more than an order of magnitude more than the fastest
uniprocessors are capable of today.\footnote{It is unlikely that the
Cpp would significantly benefit from a massively parallel
architecture.} As one can easily check, the concrete machine used by
the current Order interpreter can perform exactly $2^{65}-1$
computational steps and each step takes at least two macro
replacements. Based on these numbers, and a simple series of
divides, we can predict that if one would start the program today,
one would have to wait more than 2000 years for the failure. Of
course, the Order interpreter could easily be extended with a couple
of hundred macros to prolong the wait well beyond the estimated
lifetime of the sun\footnote{10 billion years, according to some
sources.}. The point of all this is to assure the reader that the
theoretical computational incompleteness of the Order interpreter is
the last thing to worry about. A correct Order program may
definitely fail to preprocess, but it is unlikely to happen due to
the computational incompleteness of the Order interpreter.

\input{bottles.ltx} %***********************************************

\chapter{Getting Started} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This chapter explains how to get started with experimenting with the
Order interpreter.

\section{Requirements} %********************************************

The Order interpreter, which is fully \cite{c:1999} conforming, has
been developed with the \cite{gcc} CPP version 3.x and contains no
workarounds for broken preprocessors. The Order interpreter
\emph{requires variadics and placemarkers} as defined in
\cite{c:1999}. The Wave preprocessor \cite{wave} is also known to be
able to preprocess Order. A standards conforming C preprocessor and
a fast computer with plenty of memory is recommended for
experimenting with the interpreter. Be aware that despite the
simplicity of the C preprocessor and the existence of stable
standard for years and years, many preprocessors are broken beyond
repair, either strictly not conforming to the standard or employing
macro replacement algorithms that seem to take at least exponential
time. At the time of writing, even GNU CPP, a rather highly
standards conforming\footnote{But \emph{not} 100\% conforming, at
the time of writing, due to eager elimination of whitespace
separations.} and relatively efficient preprocessor, doesn't free
memory during macro replacement, which makes the evaluation of
non-trivial C preprocessor metaprograms extremely inefficient.

\section{The Interactive Test Shell} %******************************

The interpreter comes with a simple Bash-shell script\footnote{The
script also makes use of some standard commands such as \emph{grep},
\emph{sed} and \emph{head}. You'll probably have those commands if
you have the Bash-shell.} that can be used for writing simple
regression tests and interactive evaluation of Order expressions.
The \code{test-shell} script can be found in the \code{test}
directory.

When you start the test shell by typing \code{./test-shell}, the
script first tries to find a preprocessor to use and will ask you
which preprocessor you wish to use:
\begin{verbatim}
Choose preprocessor [cpp] ?
\end{verbatim}
You must type in the name of the preprocessor you want to use.
\begin{verbatim}
Choose preprocessor [cpp] ? cpp
Wrote cpp to the .preprocessor file.
\end{verbatim}
The default is saved to a configuration file. To change the default,
you need to remove the configuration file.

It then greets you with the main menu:
\begin{verbatim}
Database: test.db
Backed up database to: test.db.bak
[(A)dd | (C)lean | (D)ebug | (M)ode: 1 | (Q)uit | (S)how | (T)est | Eval]?
\end{verbatim}
To make a selection in the menus, press a single key. The default
action can be activated with any key other than the explicitly
assigned keys for non-default actions. In the main menu above, the
\code{Eval}-action is the default action.

Most of the test shell should be quite intuitive. To run all the
tests, you can press \code{t} in the main menu, and assuming you have
a conforming preprocessor, you should see something like this:
\begin{verbatim}
Total of 360 tests.
Running positive tests...
Comparing results...
Running and comparing negative tests...
Time: 3.460 seconds
\end{verbatim}
To evaluate an expression just press enter in the main menu and
you'll get the prompt:
\begin{verbatim}
Term:
\end{verbatim}
You can now type an expression to evaluate. For example, you could
type:
\begin{verbatim}
Term: 8add(20,3)
\end{verbatim}
and then press enter to see:
\begin{verbatim}
Time: 0.082 seconds
Result: 23
\end{verbatim}
which shows that it took \verb/0.082 seconds/\footnote{Most of that
time is overhead of starting the C preprocessor.} to evaluate the
expression and the result of the expression was \code{23}.

If you make a programming error, the C preprocessor may essentially
eat all memory causing your computer to become unresponsive. In such
a case you can safely hit \code{Ctrl-C} to break the C preprocessor
and then restart the test shell. In most cases, however, a typo
simply causes garbage output. The test shell does not immediately
display the result unless it fits within one line. This helps to
avoid having to watch (hundreds of) thousands of garbage tokens
scroll through. Theoretically, the Order interpreter can generate
output whose length is counted in billions of billions of tokens.
Most likely your computer runs out of memory before that many tokens
are generated.

The debug command of the main menu is used to trace the execution of
Order programs at the \emph{continuation machine} level. It is
mostly useful for debugging the interpreter itself, but it can be
useful for the purpose of understanding how the interpreter actually
works.\footnote{The primary purpose of the test shell is to simplify
testing the interpreter.}

The mode flag in the main menu refers to the use of the built-in
\emph{debug mode} of the interpreter. When the flag is \code{1}
debug mode is active. In the debug mode, the interpreter performs
some syntax, type and precondition checks. These checks are normally
disabled, because they have a significant performance penalty, but
they can greatly help to pinpoint errors in Order programs during
development.

You should now be ready to follow the introduction to the Order
language with the help of the interactive test shell. Please do try
out expressions explained in the rest of this document using the
test shell. Of course, you can also prepare a CPP source file to try
the Order interpreter, but using the test shell for trying out
simple expressions is probably much quicker. For longer programs,
involving definitions, you will need to create CPP source files.

\chapter{Core Language} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This chapter describes the core of the Order language. The Order
interpreter can be thougth of consisting of just just one macro
\code{ORDER_PP(<exp>)}. What is interesting is the expressions that
it can evaluate.

\section{Abstraction, Application and Variables} %******************

Like in other functional programming languages, the semantics of the
Order language are rooted in the lambda calculus.

Lambda calculus is basically a tiny programming language, or a
formalization of computation, suitable for mathematical analysis.
You can find dozens of introductions to the lambda calculus on the
net (e.g. \cite{wikipedia:lambda-calculus}) and there are several
books (e.g. \cite{pierce:2002, mitchell:1996}) and articles that
contain a more thorought treatment of lambda calculus. The following
very minimal introduction describes the lambda calculus using the
syntax of the Order language.

Before we continue, it should be noted that the Order interpreter
has not been designed as a pure lambda calculus evaluator. An
evaluator for the lambda calculus should output only lambda-terms.
Although the Order interpreter can interpret \emph{closed} pure
lambda-terms, it does not output only \emph{pure} lambda-terms.
The output may, and usually does, contain \emph{closures}, which are
used internally by the interpreter during evaluation.\footnote{This
not unlike in most practical functional languages, such as Scheme
\cite{dybvig:1996,abelson:1996}.} % TBD: Explain closure

There are only three kinds of terms (or expressions) in the pure
lambda calculus as shown in the EBNF-grammar\footnote{For an
introduction to context-free grammars see \cite{aho:1986}.} snippet
below.\\
\ebnf{pure-lambda-subset}\\
In plain English, \code{8fn(v,b)} is a function of one parameter
and \code{8ap(f,x)} applies the function \code{f} to the parameter
\code{x}. You may wonder why the syntax is so terse. The reason is
that while programming in Order you will be writing lots of
\code{8fn} expressions. After a while, you will be happy that the
syntax isn't any longer. As you can see, in the Order language,
variable symbols are, by default, limited to the capital alphabetic
characters prefixed by \code{8}.

The above grammar snippet described the strict lambda calculus
subset of the Order language, where a function has exactly one
parameter. The unrestricted Order language supports the definition
of functions with an arbitrary non-zero number of parameters and
the application of a function to an arbitrary number of
parameters as well as the use of any predefined symbol as a
variable:\\
\ebnf{full-lambda-subset}\\
Above, we used an extended BNF notation.\footnote{Similar extended
BNF notation is used in \cite{friedman:2001}.} In general,
$\{x\}^{+(y)}$ indicates a sequence of one or more $x$s separated by
$y$s, i.e. strings of the form $x$, $xyx$, $xyxyx$, \ldots{} In
general, $\{x\}^*$ indicates a sequence of zero or more $x$s.

In the \emph{pure} lambda calculus, everything is a function. In
other words, there are no built-in operators or constants.
Booleans, arithmetic on natural numbers, and even lists, are
represented using functions. For example, the Church-booleans are
represented as functions:
\begin{verbatim}
#define ORDER_PP_8TRUE  ORDER_PP_FN(8fn(8T, 8fn(8F, 8T)))
#define ORDER_PP_8FALSE ORDER_PP_FN(8fn(8T, 8fn(8F, 8F)))
\end{verbatim}
\code{8TRUE} is a function that takes two parameters (one by one) and
returns the first parameter. \code{8FALSE}, on the other hand,
returns the second parameter. To compute with Church-booleans, you
apply them to functions. For example, logical conjunction can be
defined as:
\begin{verbatim}
#define ORDER_PP_8AND                    \
ORDER_PP_FN(8fn(8L, 8fn(8R,              \
                        8ap(8ap(8L, 8R), \
                            8FALSE))))
\end{verbatim}
\code{8AND} is a function that takes two parameters, \code{8L} and
\code{8R}, which are assumed to be Church-booleans. It first applies
\code{8L} to \code{8R}, i.e. \code{8ap(8L,8R)}, and then applies the
result to \code{8FALSE}, i.e. \code{8ap(8ap(8L,8R),8FALSE)}. Assuming
the parameters \code{8L} and \code{8R} were Church-booleans, the
result will be another Church-boolean. For example, the value of
the expression
\begin{verbatim}
  8ap(8ap(8AND, 8TRUE), 8FALSE)
\end{verbatim}
would be equivalent to \code{8FALSE}.

There are several different strategies for the evaluation, or
reduction, of lambda calculus terms. Most programming languages use
the \emph{call-by-value} strategy, which basically means that
parameters are evaluated before application and that no evaluation
happens inside abstractions (functional values). Another strategy is
the \emph{call-by-name} strategy, which is one form of lazy
evaluation. Like most languages, the Order language uses the
call-by-value strategy. There are three main reasons for choosing
call-by-value semantics. Call-by-value is the main strategy used
by most programming languages, including C and C++, and its
performance and other characteristics are intuitively understood by
programmers.\footnote{In contrast, according to
\cite{thompson:1999}, even experts have trouble understanding the
space and time efficiency of lazy evaluation.} While the Order
language is pure, the Order language has non-mutable side-effecting
output operations, which must be sequenced properly, which is
easiest to do in a call-by-value language. The third reason is that
the purely functional nature of the underlying C preprocessor macro
expansion mechanism makes it impossible to use any of the tricks
based on mutation used by efficient non-strict languages.




%%%%%%%%%%%%%% \footnote{In this document, \code{X ==> Y} means that \code{X} evaluates to \code{Y}.}

\subsection{Closed Terms} %=========================================

As you have already seen, the Order language implements lexical binding
and allows you to manipulate functions as first-class values. It is very
easy to create and use higher-order functions in the Order language.

When an occurrence of a symbol as a variable in an expression is
ultimately lexically surrounded by a function-abstraction with a pattern
that contains the same symbol, the occurrence of the variable is called
\emph{bound}. For example, in the expression \code{8fn(8X,8X)} the second
\code{8X} is a bound variable. If a variable is not bound it is called
\emph{free}. For example, in the expression \code{8fn(8X,8Y)} the variable
\code{8Y} is free.

The Order language only assigns semantics to terms that are closed. For
example, if you try to evaluate the expression \code{8ap(8fn(8X,8Y),8fn(8X,8X))},
the output from the interpreter is undefined.\footnote{Try to evaluate the
expression in the test shell. Try the expression in both debug and non-debug
mode. Can you count on getting the same result with the next version of the interpreter?}

\subsection{Binding} %==============================================

Even though the pure lambda calculus is Turing complete, it is not a
very practical programming language. Order, like all real
programming languages, extends the pure lambda calculus with many
syntactic forms to make programming easier and the results of
computations immediately useful.

The first extension we'll look at is the \code{8let}-expression used
for local lexical bindings. The following production describes the
syntax of the \code{8let}-derived form:\\
\ebnf{let}\\
Extending the lambda-calculus with the \code{8let}-derived form doesn't
make the language stronger. An expression of the form \code{8let(8X,i,e)}
means that the value of the expression \code{i} is bound to the variable
\code{8X} in the expression \code{e}. A \code{8let} expression of the
previous form is equivalent to an expression of the form
\code{8ap(8fn(8X,e),i)}. In other words, the \code{8let}-derived form can
be described in terms of the pure lambda calculus. The following example
simply illustrates the semantics of \code{8let} and Church-booleans:
\begin{verbatim}
8let((8T, 8fn(8T, 8F,
              8T)),
     8ap(8T, 8T, 8T)) == 8fn(8T, 8F,
                             8T)
\end{verbatim}

\section{Built-in Features} %**************************************

This section describes some of the basic built-in operators, special
forms, derived forms and constants of the Order language.

Built-in operators, in the Order language, are basically just built-in
functions of the Order language that are applied using an operator
application expression rather than a function application
expression.\footnote{You could think that a function application is really
an application of the binary operator \code{8ap}.} The following grammar
snippet describes the syntax of function application expressions:\\
\ebnf{function-application}\\
For example, the expression \code{8inc(1)} applies the unary operator
\code{8inc} to the value of the constant \code{1}.\footnote{We will get back
to the issues regarding constants and values later.}

\subsection{Conditionals} %=========================================

Like we saw earlier, it is possible to represent booleans in the pure
lambda calculus as functions. Order also provides direct support for
booleans and conditional expressions.

\subsubsection{Conditional Expressions and Boolean Constants} %---

The following grammar snippet describes the syntax for boolean constants
and conditional expressions:\\
{\small \ebnf{conditionals} }\\
The conditional special form \code{8if}, and the derived forms \code{8and} and
\code{8or} are lazy. Only the first subexpression of these forms is always
evaluated. In particular,
\begin{itemize}
\item an \code{8if} evaluates to the value of the second or third
      subexpression depending on whether the first subexpression evaluates
      to the boolean true or false value, respectively.
\item an \code{8and} evaluates to the boolean false value if the first
      subexpression evaluates to the boolean false value and otherwise it
      evaluates to the value of the second subexpression.
\item an \code{8or} evaluates to the boolean true value if the first
      subexpression evaluates to boolean true value and otherwise it
      evaluates to the value of the second subexpression.
\end{itemize}
It is important to see that only the first subexpression of the forms
\code{8if}, \code{8and} and \code{8or} needs to have a boolean value. Consider
the following example:
\begin{verbatim}
8and(8true, 8fn(8X, 8X)) == 8fn(8X, 8X)
\end{verbatim}



\chapter{Examples} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This chapter presents several Order programming examples. All of the
examples in this chapter are also available in complete source code
form. In fact, a simple script extracts the text of the examples
from the actual source code.

An attempt has been made to make each example mostly self-contained
at the cost of repeating some explanations. A reader having a good
understanding of programming should be able to follow the examples
with a minimal understanding of the Order language. The narrative
body of each example briefly motivates the example and gives a bit
of background information, but doesn't aim to explain every detail
of each Order language feature used.

Most of the examples are meant to apply programming techniques that
are likely to be useful in practice to the kind of problems that one
might encounter in practice, but none of the examples is meant to be
an industry strength solution. This is emphasized by exercises,
given at the end of each example, that ask the reader to modify and
extend the examples in various ways. While it is probably too much
to ask the reader to complete all the exercises, all readers are
strongly recommended to preprocess one or more examples to actually
see what is ultimately being generated.

\input{fibonacci.ltx}
\input{exp-delay.ltx}
\input{average.ltx}
\input{binary.ltx}
\input{duffs-device.ltx}
\input{array-ops.ltx}
\input{catch-builtin.ltx}
\input{is-function.ltx}
\input{is-integral.ltx}
\input{streams.ltx}

\appendix %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{The Order Grammar} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

{\small \label{grammar-summary} \ebnf{grammar-summary} }

% Bibliography %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\bibliographystyle{alpha}
\bibliography{bibliography}
\addcontentsline{toc}{chapter}{Bibliography}

\end{document}
